---
title: 'COVID-19 From Pandemic to Endemic Transition in Indonesia: A Review'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
A Short Analysis by: Elkan Jeremy Ishak

# Introduction
It's been two years since the WHO officially declared COVID-19 as a global pandemic and countries still battling this never ending disease. World's economy has been
impacted for the last two years, including Indonesia. Business disrupted. Lay-off everywhere. One question arises: When will it end?

Now, as the latest wave of infections driven by the fast-spreading Omicron variant rapidly subsides, many are beginning to question:
Is COVID-19 becoming endemic?

Indonesia's closest neighbors have declared that a "transition to endemic" phase will be implemented. Malaysia, starting 1 April 2022, will begin the transition.
Followed with Thailand in 1 July 2022. How about Indonesia? Are we ready for this life-changing challenge? In this short analysis, I'm going to present about
Indonesia's readiness for implementing the transition of pandemic.

There are three key aspects that will be explored and discussed:
-Vaccination effects on Case Fatality Rate (CFR) in Indonesia
-Severity of Delta vs Omicron variant
-Indonesia's CFR comparison with neighboring countries

Notice that CFR is mentioned many times. In short, CFR is a parameter to measure of how 'severe' is a disease by comparing the total of death caused by an infection 
with the total cases of an infection. Details will be presented later on.

# Environment Setup, Data Retrieval, and Data Processing
## Import Libraries
```{r Import libraries, message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr) 
library(readr)
library(tidyr)
library(ggplot2)
library(plotly)
```
Dataset is retrieved from the COVID-19 data from John Hopkins University.

URL: https://www.kaggle.com/datasets/antgoldbloom/covid19-data-from-john-hopkins-university

**Updated on: 21 March 2022**

PS: All scripts in this document are designed to handle any future dataset updates from the URL above. Replace files in the './Dataset/' path with newest downloaded files from the URL.

## Import Datasets
```{r Import Dataset, warning=FALSE}
confirmed <- read_csv('Dataset/CONVENIENT_global_confirmed_cases.csv') #Total recorded COVID-19 cases from day to day in every country
deaths <- read_csv('Dataset/CONVENIENT_global_deaths.csv') #Total recorded COVID-19 deaths from day to day in every country
vaccine <- read_csv('Dataset/vaccine.csv')
```
```{r Confirmed case data summary, eval=FALSE, include=FALSE}
str(confirmed)
colnames(confirmed)
head(confirmed)
```
```{r Confirmed death data summary, eval=FALSE, include=FALSE}
str(deaths)
colnames(deaths)
head(deaths)

```
## Datasets Cleaning
Dataset cleaning task for `confirmed` and `deaths`:

- It is found that Australia, Canada, China, Denmark, France, Netherlands, New Zealand, and United Kingdom each has more than 1 column
(each province/region/state are separated in the dataset) and R automatically rename their column names. Specific data on each province/
region are not needed. Hence, we need first calculate the sum cases of those countries.
- 'Province/State' row is not needed.
- Rename 'Country/Region' column name to 'Time'
- Convert 'Time' column (which contains date) from character to date datatype.
- Countries name are a variable which should be in one column. `pivot_longer` should be done to compact countries name into one column.

```{r Confirmed case data cleaning processes}
confirmed <- confirmed %>%
  filter(`Country/Region` != 'Province/State') %>% #Remove 'Province/State row'
  mutate(`Country/Region` = as.Date(`Country/Region`, format='%m/%d/%y')) %>% #Convert to Date format
  rename(Time = `Country/Region`) %>%
  mutate_if(is.character, as.numeric) %>% #convert all columns (except Time) to numeric
  rowwise() %>%
  mutate(Australia = sum(across(starts_with("Australia")))) %>% #Sum Australia...11, Australia...12, ..., Australia...18 to Australia
  mutate(Canada = sum(across(starts_with("Canada")))) %>%
  mutate(Denmark = sum(across(starts_with("Denmark")))) %>%
  mutate(China = sum(across(starts_with("China")))) %>%
  mutate(France = sum(across(starts_with("France")))) %>%
  mutate(Netherlands = sum(across(starts_with("Netherlands")))) %>%
  mutate(`New Zealand` = sum(across(starts_with("New Zealand")))) %>%
  mutate(`United Kingdom` = sum(across(starts_with("United Kingdom")))) %>%
  select(-(Australia...11:Australia...18)) %>%
  select(-(Canada...42:Canada...57)) %>%
  select(-(China...61:China...94)) %>%
  select(-(Denmark...105:Denmark...107)) %>%
  select(-(France...122:France...133)) %>%
  select(-(Netherlands...196:Netherlands...200)) %>%
  select(-(`New Zealand...201`:`New Zealand...202`)) %>%
  select(-(`United Kingdom...262`:`United Kingdom...275`)) %>%
  pivot_longer(!Time, names_to='Country', values_to='ConfirmedCase') %>% #Compact data
  arrange(Country) #Sort by country's name
```
```{r Post-cleaning confirmed case data check}
head(confirmed)
```
```{r Confirmed death data cleaning processes}
deaths <- deaths %>%
  filter(`Country/Region` != 'Province/State') %>% #Remove 'Province/State row'
  mutate(`Country/Region` = as.Date(`Country/Region`, format='%m/%d/%y')) %>% #Convert to Date format
  rename(Time = `Country/Region`) %>%
  mutate_if(is.character, as.numeric) %>% #convert all columns (except Time) to numeric
  rowwise() %>%
  mutate(Australia = sum(across(starts_with("Australia")))) %>% #Sum Australia...11, Australia...12, ..., Australia...18 to Australia
  mutate(Canada = sum(across(starts_with("Canada")))) %>%
  mutate(Denmark = sum(across(starts_with("Denmark")))) %>%
  mutate(China = sum(across(starts_with("China")))) %>%
  mutate(France = sum(across(starts_with("France")))) %>%
  mutate(Netherlands = sum(across(starts_with("Netherlands")))) %>%
  mutate(`New Zealand` = sum(across(starts_with("New Zealand")))) %>%
  mutate(`United Kingdom` = sum(across(starts_with("United Kingdom")))) %>%
  select(-(Australia...11:Australia...18)) %>%
  select(-(Canada...42:Canada...57)) %>%
  select(-(China...61:China...94)) %>%
  select(-(Denmark...105:Denmark...107)) %>%
  select(-(France...122:France...133)) %>%
  select(-(Netherlands...196:Netherlands...200)) %>%
  select(-(`New Zealand...201`:`New Zealand...202`)) %>%
  select(-(`United Kingdom...262`:`United Kingdom...275`)) %>%
  pivot_longer(!Time, names_to='Country', values_to='ConfirmedDeath') %>% #Compact data
  arrange(Country) #Sort by country's name
```
```{r Post-cleaning confirmed death data check}
head(deaths)
```
```{r Vaccination data cleaning processes}
vaccine <- vaccine %>%
  mutate(date = as.Date(date, format='%m/%d/%Y')) %>%
  rename(c("Time" = "date", "Country" = "location", "FullyVaccinatedPercent" = "people_fully_vaccinated_per_hundred")) 
```
```{r Post-cleaning vaccination data check}
head(vaccine)
```

## Joining Confirmed Case, Death, and Vaccine into one dataframe
```{r}
df <- confirmed %>%
  inner_join(deaths, by=c('Time','Country')) %>%
  left_join(vaccine, by=c('Time','Country'))
```
```{r}
head(df)
```

**Final dataframe has been established. Hence, we can proceed to the data exploration step.**

# Analysis of Vaccination Effect on Indonesia's Case Fatality Rate
## Case Fatality Rate
Case Fatality Rate is a parameter that measures the severity of the disease that causes death. CFR is the proportion of people diagnosed with a certain disease, who end up dying of it. For example, among a total of 200 patients with disease A, 20 of them died from the same disease within 30 days. Thus,

The 30-day case fatality rate = $\frac{20}{200} \times 100\% = 10\%$

## CFR Calculation
Actually, CFR is not designed to be calculated on an ongoing pandemic. Ideally, CFR should be calculated after the end of a pandemic, because there will be a bias. From the CFR definition above, CFR is calculated by simply dividing death caused by a disease with the total confirmed cases. At glance, we are tempted to simply divide the total death with the total confirmed new cases in each days. But, one should be noted that people are not die instantly from a disease. A person whom die by a COVID-19 had been infected in around 10 days before. Thus, there is a 'delay' bias. We will explore this bias and how to ease this bias.

First, we need to find the delay. This are done by plotting ConfirmedCase and ConfirmedDeath to find each 'peak'.
```{r Select data for Indonesia}
indonesia <- df %>%
  filter(Country == 'Indonesia')
```
```{r Indonesia summary}
tail(indonesia)
```
### Making the dataframe for plotting on ggplot
```{r Making dataframe for plotting on ggplot}
indonesia_for_ggplot <- indonesia %>%
  pivot_longer(cols=c(ConfirmedCase, ConfirmedDeath, FullyVaccinatedPercent), names_to='Type', values_to='Values')
```
```{r}
tail(indonesia_for_ggplot)
```
### Plotting Indonesia's Confirmed Case, Death vs Time
```{r Plot of Indonesia\'s Confirmed Cases, Confirmed Death, and Vaccination Rate}
ggplot(indonesia_for_ggplot, aes(x=Time, y=Values, colour=Type, group=Type)) +
  geom_line() +
  facet_grid(Type ~ ., scales = 'free_y') +
  labs(title='Indonesia\'s COVID-19 Confirmed Case and Confirmed Death')
```
From the presented plot above, it can be seen that there are 3 waves of COVID-19 in Indonesia. If we take a look at Wave 2, which is happened around July 2021, it is obvious that there is a unnoticeable interval between the `ConfirmedCase` and `ConfirmedDeath` peak. If we look closer, the peak of `ConfirmedCase` happened on 15 July 2021. However, the peak of `ConfirmedDeath` happened on 27 July 2021. We can conclude that there are 12 days of delay in Indonesia. Thus, to estimate the CFR on an ongoing COVID-19, we need to shift the `ConfirmedDeath` by 12 days.

>Reference: Thomas, B. S., & Marks, N. A. (2021). Estimating the case fatality ratio for COVID-19 using a time-shifted distribution analysis. Epidemiology and infection, 149, e197. https://doi.org/10.1017/S0950268821001436

### Dataframe with Shifted Death
```{r Creating a new dataframe for shifting death}
indonesia_death_shifted <- indonesia %>%
  select(Time, ConfirmedDeath) %>%
  mutate(Time = Time - 12) %>% #Shift Date by 12 days before
  rename(ConfirmedDeathShifted = ConfirmedDeath)
```
```{r}
tail(indonesia_death_shifted)
```
```{r Joining Shifted dataframe with main df}
indonesia_final_cfr <- indonesia %>%
  inner_join(indonesia_death_shifted)
```
```{r}
tail(indonesia_final_cfr)
```
Lastly, we created a dataframe to calculate CFR on each day.
## Final Indonesia's CFR data
```{r}
indonesia_final_cfr <- indonesia_final_cfr %>%
  mutate(CFR = ConfirmedDeathShifted/ConfirmedCase*100) %>% #CFR Calculation
  select(-ConfirmedDeath) %>%
  filter(ConfirmedCase > 200)
indonesia_final_cfr[is.na(indonesia_final_cfr)] = 0 #Make all NaN in the CFR column to 0
```
```{r}
tail(indonesia_final_cfr)
```
## Dataframe for plotting
```{r}
indonesia_for_cfr_vaccine_plot <- indonesia_final_cfr %>%
  pivot_longer(cols = c(CFR, FullyVaccinatedPercent), names_to='Type', values_to='Value')
```
```{r}
head(indonesia_for_cfr_vaccine_plot)
```
## Final plot of CFR and Vaccination Rate vs Time
```{r}
vaccine_effect <- ggplot(indonesia_for_cfr_vaccine_plot, aes(x=Time, y=Value)) + facet_grid(Type~., scales = "free_y") +
  geom_smooth(data=subset(indonesia_for_cfr_vaccine_plot, Type == 'CFR')) +
  geom_line(data=subset(indonesia_for_cfr_vaccine_plot, Type == 'FullyVaccinatedPercent'), size=1, color='green') +
  labs(title='Effect of Vaccination on Case Fatality Rate in Indonesia', x='Date', y='Value (%)') +
  theme(
     plot.background = element_rect(fill = "#3d989b",colour = "#3d989b"),
     # gets rid of white border around plot: 
     panel.border = element_blank(),
     plot.title = element_text(colour="#3b365f", face='bold'),
     panel.background = element_rect(fill='#faede4')
   )
vaccine_effect
```
From the plot above, it can be seen that vaccination give a significant effect in decreasing the CFR in Indonesia. With vaccination, severity of COVID-19 is lowered. Vaccination program held by Indonesia's government is proven to be successful. 

# Delta and Omicron Severity Comparison
Many claims that Omicron is less severe compared to the Delta variant. In the following section, we will prove by their CFR compared to each other. From the data, we know that Delta variant is the most dominant variant from 9 August 2021 to 13 December 2021 in Indonesia. And there were a transition period from the Delta variant to Omicron between 13 December 2021 and 7 February 2022. Since 7 February 2022, Omicron has been regarded as the most dominant variant in Indonesia.

We will explore the distribution of CFR for each variant.

Data Source: https://ourworldindata.org/grapher/covid-variants-bar?country=~IDN

## Dataframe preparation
```{r}
delta <- indonesia_final_cfr %>%
  filter(Time > '2021-08-09') %>%
  filter(Time < '2021-12-13')
delta$Variant <- "Delta"

omicron <- indonesia_final_cfr %>%
  filter(Time > '2022-02-07')
omicron$Variant <- "Omicron"

indonesia_variant <- delta %>%
  full_join(omicron)
```
## Plot with box plot
```{r}
ggplot(indonesia_variant, aes(Variant, CFR)) +
  geom_boxplot(color = '#3b365f', fill='#a4ebe1') +
  labs(title="Delta vs Omicron Case Fatality Rate Distribution") +
  theme(
     plot.background = element_rect(fill = "#a4ebe1",colour = "#a4ebe1"),
     # gets rid of white border around plot: 
     panel.border = element_blank(),
     plot.title = element_text(colour="#3b365f", face='bold'),
     panel.background = element_rect(fill='#faede4')
   )
```
## Hypothesis testing with t-test
```{r Two Sample T-Test}
t_test <- t.test(delta$CFR, omicron$CFR)
t_test
```
From the plot above, it can be seen that the Omicron variant has much lower CFR compared to the Delta variant. With t-test, it is also concluded that with 95% confidence level, the probability that Omicron variant is not less severe that the Delta variant is $2 \times 10^{-16}$. Thus, we can conclude that the Omicron variant is proven statistically that less severe compared to the Delta variant.

# Indonesia's Situation compared with Malaysia and Thailand
In the following section, CFR of Malaysia and Thailand will be compared to Indonesia's CFR. Since Malaysia and Thailand are going to treat COVID-19 as an endemic, this section will predict Indonesia's situation compared to those countries. Steps are similar with analysing Indonesia's CFR, where the death data shifting will be done.
```{r Selecting data for Malaysia}
malaysia <- df %>%
  filter(Country == 'Malaysia')
```
```{r Malaysia Shifting Death}
#new death dataframe
malaysia_death_shifted <- malaysia %>%
  select(Time, ConfirmedDeath) %>%
  mutate(Time = Time - 16) %>% #Shift Date by 16 days before
  rename(ConfirmedDeathShifted = ConfirmedDeath)
tail(malaysia_death_shifted)
```
```{r Joining Shifted dataframe with main df malay}
malaysia_final_cfr <- malaysia %>%
  inner_join(malaysia_death_shifted)
tail(malaysia_final_cfr)
```

```{r}
malaysia_final_cfr <- malaysia_final_cfr %>%
  mutate(CFR = ConfirmedDeathShifted/ConfirmedCase*100) %>% #CFR Calculation
  select(-ConfirmedDeath) %>%
  filter(ConfirmedCase > 50)
malaysia_final_cfr[is.na(malaysia_final_cfr)] = 0 #Make all NaN in the CFR column to 0
tail(malaysia_final_cfr)
```
```{r}
ggplot(malaysia_final_cfr, aes(Time, CFR)) +
  geom_point() +
  geom_smooth()
```
```{r Selecting data for Thai}
thai <- df %>%
  filter(Country == 'Thailand')
head(thai)
```
```{r thai Shifting Death}
#new death dataframe
thai_death_shifted <- thai %>%
  select(Time, ConfirmedDeath) %>%
  mutate(Time = Time - 5) %>% #Shift Date by 5 days before
  rename(ConfirmedDeathShifted = ConfirmedDeath)
tail(thai_death_shifted)
```
```{r Joining Shifted dataframe with main dataframe}
thai_final_cfr <- thai %>%
  inner_join(malaysia_death_shifted)
tail(thai_final_cfr)
```

```{r}
thai_final_cfr <- thai_final_cfr %>%
  mutate(CFR = ConfirmedDeathShifted/ConfirmedCase*100) %>% #CFR Calculation
  select(-ConfirmedDeath) %>%
  filter(ConfirmedCase > 30)
thai_final_cfr[is.na(thai_final_cfr)] = 0 #Make all NaN in the CFR column to 0
tail(thai_final_cfr)
```
```{r}
ggplot(thai_final_cfr, aes(Time, CFR)) +
  geom_point() +
  geom_smooth()
```
```{r}
indo_malay_thai_cfr <- indonesia_final_cfr %>%
  full_join(malaysia_final_cfr) %>%
  full_join(thai_final_cfr)
tail(indo_malay_thai_cfr)
```

```{r}
ggplot(indo_malay_thai_cfr, aes(Time, CFR, group = Country, color=Country)) +
  geom_smooth() + 
  labs(title='COVID-19 Case Fatality Rate in Indonesia, Malaysia, and Thailand') +
  theme(
     plot.background = element_rect(fill = "#3d989b",colour = "#3d989b"),
     # gets rid of white border around plot: 
     panel.border = element_blank(),
     plot.title = element_text(colour="#3b365f", face='bold'),
     panel.background = element_rect(fill='#faede4')
   )
```
It can be seen that now Indonesia's CFR lies in similar value with Malaysia and Thailand's CFR. The severity of COVID-19 in Indonesia should be the same as Malaysia and Thailand.

# Conclusions
1. Indonesia vaccination program has succeeded to suppress the severity of COVID-19 in Indonesia. It is proven by the decrease of CFR and increasing vaccination rate in Indonesia. The decreasing CFR value is also contributed by the domination of Omicron variant in Indonesia. Vaccination is the key aspect to endemic transition. Mutation to Omicron variant is also prove that the virus exists not to infect people severely, but only to retain its existence.
2. Comparing to Malaysia and Thailand, Indonesia has similar COVID-19 situation in terms of severity. One can assume that Indonesia has similar vaccination and healthcare condition compared to Malaysia and Thailand.
3. It is still not safe to say that Indonesia is ready for the transition, because this short analysis only covers from the terms of COVID-19's severity and vaccination success. But, it is worth for the Government to do more research of the feasibility. Economics and social-politic aspects haven't covered in this analysis.